---
title: "Juvenile Brown Trout Survival & Emigration Dynamics"
subtitle: "Upper Clark Fork River — Mark-Recapture Power Analysis"
date: today
date-format: "MMMM D, YYYY"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(cowplot)
library(kableExtra)
library(scales)

# True simulation parameter values — must match power_analysis_CFBrown_loop.R
n_occ            <- 5
psi_main_down_true <- 0.04
psi_main_back_true <- 0.04

# Scenario reference table
scenarios <- tibble(
  scenario     = c(1, 2, 3, 5, 6, 7, 8, 10),
  n_tagged     = c(500, 1000, 2000, 1000, 1000, 1000, 1000, 1000),
  phi_annual   = c(0.20, 0.20, 0.20, 0.10, 0.30, 0.20, 0.20, 0.10),
  p_efish_true = c(0.25, 0.25, 0.25, 0.25, 0.25, 0.15, 0.25, 0.25),
  use_inform   = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE),
  label        = c(
    "500 tagged",  "1000 tagged", "2000 tagged",
    "Low survival\n(10%)",  "Higher survival\n(30%)",  "Poor efish\n(p = 0.15)",
    "1000 tagged\n(vague priors)", "Low survival\n(vague priors)"
  ),
  label_short = c(
    "500", "1000", "2000",
    "Low φ", "High φ", "Low p_efish",
    "1000 (vague)", "Low φ (vague)"
  )
) |>
  mutate(phi_occ = phi_annual^(1 / n_occ))

# Load simulation results if available
results_file     <- "simulation_output_CFBrown.rds"
results_available <- file.exists(results_file)

if (results_available) {
  all_data <- read_rds(results_file) |>
    left_join(scenarios, by = c("scenario", "n_tagged", "use_inform")) |>
    mutate(
      bias_phi_main  = phi_main_mean  - phi_occ,
      bias_phi_trib  = phi_trib_mean  - phi_occ,
      bias_psi_down  = psi_down_mean  - psi_main_down_true,
      bias_psi_back  = psi_back_mean  - psi_main_back_true,
      bias_p_gem     = p_gem_mean     - 0.76
    )
}

# Shared plot theme
theme_cf <- function() {
  theme_cowplot(font_size = 12) +
    theme(strip.background = element_rect(fill = "grey92"),
          legend.position  = "bottom")
}
```

## Overview

This document presents a simulation-based power analysis for a multi-year juvenile
brown trout (*Salmo trutta*) survival study on the Upper Clark Fork River, Montana.
The study uses PIT (Passive Integrated Transponder) tagging combined with stationary
antenna arrays and electrofishing recapture surveys to estimate reach-specific
survival and emigration dynamics.

**Core scientific questions:**

1. Is survival of juvenile brown trout in the mainstem low enough to be a
   demographic bottleneck?
2. Are fish emigrating from tributaries permanently leaving the upper river, or do
   they remain in the mainstem and contribute to the resident population?

The power analysis asks: *how many fish need to be tagged to answer these questions
with acceptable precision?*

---

## Study System

### Spatial Layout

The monitoring network spans approximately 22 river miles of the upper Clark Fork,
anchored by three mainstem PIT antenna arrays and three tributary mouth antennas.

```{mermaid}
%%| fig-cap: "Study system schematic. Boxes represent states in the statistical model; arrows show allowed transitions with the antenna that fires on each crossing."
%%| fig-width: 9

flowchart TD
    T["**State 1**\nAlive in Tributary\n*(tribs 1, 2, 3 lumped)*"]
    M["**State 2**\nAlive in Mainstem\nWarm Springs – Gembeck Rd\n*(primary study area)*"]
    L["**State 3**\nAlive in Mainstem\nGembeck Rd – Sager Ln\n*(secondary study area)*"]
    PS["**State 4**\nEmigrated past Sager Ln\n*(absorbing)*"]
    PW["**State 5**\nEmigrated past Warm Springs\n*(absorbing)*"]
    D["**State 6**\nDead\n*(absorbing)*"]

    T -- "Trib antenna\n*p_trib ≈ 0.93*" --> M
    M -- "Trib antenna\n*p_trib ≈ 0.93*" --> T
    M -- "Gembeck antenna\n*p_gem ≈ 0.76*" --> L
    L -- "Sager antenna\n*p_sag ≈ 0.76*" --> PS
    M -- "WS antenna\n*p_ws ≈ 0.76*" --> PW
    T --> D
    M --> D
    L --> D

    style T  fill:#d4ecd4,stroke:#4a7c4a
    style M  fill:#d0e4f7,stroke:#3a6ea8
    style L  fill:#e8d8f0,stroke:#7a4a9a
    style PS fill:#f5e6c8,stroke:#a07030
    style PW fill:#f5e6c8,stroke:#a07030
    style D  fill:#f0d0d0,stroke:#a03030
```

### Antenna Design and Detection Efficiency

All antenna detections are **transition-based**: an antenna fires only when a fish
physically crosses that boundary. This is a critical design feature — the
Gembeck + Sager antenna pair constitutes a **double-observer system** for
downstream emigrants, allowing detection efficiency at Gembeck to be estimated
from the data rather than assumed.

```{r}
#| label: tbl-antennas
#| tbl-cap: "Monitoring infrastructure and pilot-estimated detection probabilities."
#| echo: false

tibble(
  Location          = c("Trib 1, 2, 3 mouth antennas", "Warm Springs (mainstem)",
                        "Gembeck Rd (mainstem)", "Sager Ln (mainstem)",
                        "Electrofishing (mainstem)"),
  Type              = c("Passive PIT", "Passive PIT", "Passive PIT",
                        "Passive PIT", "Active recapture"),
  `Detection model` = c("Transition: states 1 ↔ 2", "Transition: state 2 → 5",
                        "Transition: state 2 → 3", "Transition: state 3 → 4",
                        "Proximity: state 2"),
  `Pilot estimate`  = c("0.93", "0.76 (assumed)", "0.76", "0.76 (assumed)", "0.15 – 0.35")
) |>
  kbl(align = c("l", "l", "l", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(0, bold = TRUE)
```

---

## Statistical Model

### Hidden Markov Model Structure

Fish fate is modeled as a Hidden Markov Model (HMM) in which the true state
$z_{i,t}$ (alive in a given reach, or dead) evolves according to a Markov
transition matrix and is observed imperfectly through the antenna and
electrofishing detection process.

At each occasion $t$, the state of fish $i$ evolves as:

$$z_{i,t} \sim \text{Categorical}(\mathbf{T}_{z_{i,t-1},\cdot})$$

where $\mathbf{T}$ is the $6 \times 6$ transition matrix parameterized by
reach-specific survival ($\phi$) and movement probabilities ($\psi$). The
initial state $z_{i,1}$ is known from the tagging location.

### Transition Matrix

Movement from the primary study area (state 2) uses a Dirichlet parameterization
to ensure that the three directional probabilities (return to trib, downstream,
upstream) sum to $\leq 1$:

$$\psi_{\text{back}}, \psi_{\text{down}}, \psi_{\text{up}} = \psi_{\text{move}} \times \text{Dir}(\alpha_1, \alpha_2, \alpha_3)$$

### Detection Priors

Detection probability priors are **logit-normal**, centered on pilot estimates
when informative priors are requested:

$$\text{logit}(p) \sim \mathcal{N}(\mu, \sigma^2)$$

| Prior type | $\mu$ | $\sigma$ (logit scale) |
|---|---|---|
| Informative | $\text{logit}(\hat{p}_\text{pilot})$ | 0.30 |
| Vague | 0 | 5.00 |

Toggling between prior types requires only a change in the data passed to JAGS —
the model file itself is unchanged.

### JAGS Model

```{r}
#| label: jags-model
#| eval: false
#| echo: true
#| code-fold: false
#| file: survival_CFBrown.txt
```

---

## Simulation Design

### Approach

The power analysis uses a **parametric bootstrap** design:

1. Simulate a dataset under known true parameters
2. Fit the JAGS model to the simulated data
3. Record posterior means and SDs for key parameters
4. Repeat 25 times per scenario
5. Summarize bias, precision (posterior SD), and RMSE across replications

Power is assessed by whether the posterior SD is small enough to distinguish
biologically meaningful differences — not a formal hypothesis test.

### True Parameter Values

```{r}
#| label: tbl-params
#| tbl-cap: "True parameter values used in all simulations (baseline survival scenario). Annual survival is converted to per-occasion survival as φ^(1/5)."
#| echo: false

tibble(
  Parameter = c(
    "φ_trib, φ_main, φ_lower (per occasion, baseline)",
    "ψ_trib→main (emigrate from trib per occasion)",
    "ψ_main→trib (return to trib per occasion)",
    "ψ_main→downstream (past Gembeck per occasion)",
    "ψ_main→upstream (past WS per occasion)",
    "ψ_sager (pass Sager given alive in state 3)",
    "p_trib (trib antenna)",
    "p_gem (Gembeck antenna)",
    "p_ws, p_sag (WS, Sager antennas)",
    "p_efish (electrofishing, baseline)"
  ),
  `True value` = c(
    "0.20^(1/5) = 0.725",
    "0.06",  "0.04",  "0.04", "0.01",  "0.40",
    "0.93",  "0.76",  "0.76", "0.25"
  ),
  Source = c(
    "Colleague email (10–30% annual range)",
    "Derived from 11–15% total emigration",
    "Professional judgment",
    "Derived from 11–15% total emigration",
    "Low; WS antenna as upstream sentinel",
    "Expected high for downstream section",
    "Pilot antenna data (2024–2025)",
    "Pilot antenna data (2024–2025)",
    "Assumed equal to Gembeck estimate",
    "Pilot electrofishing (2024–2025)"
  )
) |>
  kbl(align = c("l", "c", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) |>
  row_spec(0, bold = TRUE)
```

### Scenarios

```{r}
#| label: tbl-scenarios
#| tbl-cap: "Simulation scenarios. Scenarios 1–7 use informative detection priors from pilot data. Scenarios 8 and 10 repeat key conditions with vague priors to quantify the value of the pilot work."
#| echo: false

scenarios |>
  select(scenario, n_tagged, phi_annual, p_efish_true, use_inform) |>
  mutate(
    phi_occ    = round(phi_annual^(1/5), 3),
    use_inform = if_else(use_inform, "Informative", "Vague")
  ) |>
  rename(
    Scenario      = scenario,
    `N tagged`    = n_tagged,
    `φ (annual)`  = phi_annual,
    `φ (per occ)` = phi_occ,
    `p_efish`     = p_efish_true,
    `Det. priors` = use_inform
  ) |>
  kbl(align = c("c", "r", "c", "c", "c", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(0, bold = TRUE) |>
  pack_rows("Vary sample size (baseline survival)", 1, 3) |>
  pack_rows("Vary survival & detection (1000 fish)", 4, 6) |>
  pack_rows("Vague priors — value of pilot data", 7, 8)
```

---

## Results

```{r}
#| label: results-check
#| echo: false
#| results: asis

if (!results_available) {
  cat("> **Results pending.** The simulation output file
  (`simulation_output_CFBrown.rds`) has not yet been committed to the
  repository. Once the overnight JAGS run completes and the output is pushed
  to GitHub, this section will render automatically on the next Actions run.")
}
```

```{r}
#| label: convergence-summary
#| echo: false
#| eval: !expr results_available
#| tbl-cap: "Convergence summary by scenario (Rhat < 1.1 for all monitored parameters)."

conv_summary <- all_data |>
  group_by(scenario, label_short, n_tagged, phi_annual, use_inform) |>
  summarise(
    n_reps      = n(),
    n_converged = sum(converged, na.rm = TRUE),
    conv_pct    = round(100 * mean(converged, na.rm = TRUE)),
    .groups     = "drop"
  )

conv_summary |>
  select(Scenario = scenario, Label = label_short,
         `N tagged` = n_tagged, `φ annual` = phi_annual,
         `Informative priors` = use_inform,
         `Reps converged` = n_converged,
         `% converged` = conv_pct) |>
  mutate(`Informative priors` = if_else(`Informative priors`, "Yes", "No")) |>
  kbl(align = c("c","l","r","c","c","r","r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(0, bold = TRUE)
```

### Precision by Sample Size

Posterior standard deviation (SD) measures how precisely each parameter is
estimated. Smaller SD = tighter credible intervals = more confident conclusions.

```{r}
#| label: fig-precision-ntag
#| fig-cap: "Posterior SD for primary parameters across sample size scenarios (scenarios 1–3, informative priors, 20% annual survival). Horizontal blue lines show the true parameter value. Lower SD indicates better precision."
#| eval: !expr results_available

plot_dat <- all_data |>
  filter(scenario %in% c(1, 2, 3)) |>
  select(scenario, n_tagged, label_short,
         `Trib survival\n(φ_trib)`       = phi_trib_sd,
         `Mainstem survival\n(φ_main)`   = phi_main_sd,
         `Downstream emigration\n(ψ_down)` = psi_down_sd,
         `Gembeck detection\n(p_gem)`    = p_gem_sd) |>
  pivot_longer(-(scenario:label_short), names_to = "parameter", values_to = "post_sd") |>
  mutate(n_tagged = factor(n_tagged))

ggplot(plot_dat, aes(x = n_tagged, y = post_sd)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 0.8) +
  facet_wrap(~parameter, scales = "free_y", ncol = 2) +
  labs(x = "Number of fish tagged", y = "Posterior SD") +
  theme_cf()
```

### Bias Assessment

Bias is the mean difference between the estimated posterior mean and the known
true value. Values near zero indicate the model recovers the truth without
systematic over- or under-estimation.

```{r}
#| label: fig-bias
#| fig-cap: "Bias (estimated − true) for survival and downstream emigration across sample size scenarios. Dashed line at zero = no bias."
#| eval: !expr results_available

bias_dat <- all_data |>
  filter(scenario %in% c(1, 2, 3)) |>
  select(scenario, n_tagged,
         `Mainstem survival\n(φ_main)`     = bias_phi_main,
         `Trib survival\n(φ_trib)`          = bias_phi_trib,
         `Downstream emigration\n(ψ_down)`  = bias_psi_down) |>
  pivot_longer(-(scenario:n_tagged), names_to = "parameter", values_to = "bias") |>
  mutate(n_tagged = factor(n_tagged))

ggplot(bias_dat, aes(x = n_tagged, y = bias)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_boxplot(fill = "coral", alpha = 0.6, outlier.size = 0.8) +
  facet_wrap(~parameter, scales = "free_y", ncol = 3) +
  labs(x = "Number of fish tagged", y = "Bias (estimated − true)") +
  theme_cf()
```

### Value of Pilot Detection Data (Prior Sensitivity)

Scenarios 8 and 10 repeat baseline conditions with **vague** priors on detection
probabilities. Comparing these against matched informative-prior scenarios
quantifies how much the 2023–2025 pilot antenna and electrofishing data are
worth in terms of parameter precision.

```{r}
#| label: fig-prior-sensitivity
#| fig-cap: "Effect of prior type on posterior SD for key parameters. Each pair of boxplots shares the same n_tagged and true parameter values. 'Informative' priors use pilot detection estimates; 'Vague' priors are essentially flat."
#| eval: !expr results_available

prior_dat <- all_data |>
  filter(scenario %in% c(2, 5, 8, 10)) |>
  mutate(
    comparison = case_when(
      scenario %in% c(2, 8)  ~ "1000 fish\n20% survival",
      scenario %in% c(5, 10) ~ "1000 fish\n10% survival"
    ),
    prior_label = if_else(use_inform, "Informative\npriors", "Vague\npriors")
  ) |>
  select(comparison, prior_label,
         `φ_main SD`  = phi_main_sd,
         `ψ_down SD`  = psi_down_sd,
         `p_gem SD`   = p_gem_sd) |>
  pivot_longer(-(comparison:prior_label), names_to = "parameter", values_to = "post_sd")

ggplot(prior_dat, aes(x = prior_label, y = post_sd, fill = prior_label)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_grid(parameter ~ comparison, scales = "free_y") +
  scale_fill_manual(values = c("Informative\npriors" = "steelblue",
                                "Vague\npriors"       = "coral"),
                    guide = "none") +
  labs(x = NULL, y = "Posterior SD") +
  theme_cf()
```

### Summary Statistics

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of estimation performance across all scenarios. Mean SE = mean posterior SD across replications (precision); RMSE = root mean squared error (combined accuracy + precision); Bias = mean(estimated − true)."
#| eval: !expr results_available

all_data |>
  group_by(Scenario = scenario, Label = label_short) |>
  summarise(
    `N reps`        = sum(!is.na(phi_main_mean)),
    `φ_main bias`   = round(mean(bias_phi_main,  na.rm = TRUE), 4),
    `φ_main mean SE`= round(mean(phi_main_sd,    na.rm = TRUE), 4),
    `φ_main RMSE`   = round(sqrt(mean(bias_phi_main^2, na.rm = TRUE)), 4),
    `ψ_down bias`   = round(mean(bias_psi_down,  na.rm = TRUE), 4),
    `ψ_down mean SE`= round(mean(psi_down_sd,    na.rm = TRUE), 4),
    `p_gem mean SE` = round(mean(p_gem_sd,        na.rm = TRUE), 4),
    `% converged`   = round(100 * mean(converged, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  kbl(align = c("c", "l", rep("r", 8))) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE)
```

---

## Recommendations

```{r}
#| label: recommendations
#| echo: false
#| results: asis

if (results_available) {

  # Pull key precision numbers for the inline text
  baseline <- all_data |>
    filter(scenario == 2) |>
    summarise(
      phi_main_se  = round(mean(phi_main_sd,  na.rm = TRUE), 3),
      psi_down_se  = round(mean(psi_down_sd,  na.rm = TRUE), 3),
      conv_pct     = round(100 * mean(converged, na.rm = TRUE))
    )

  cat(glue::glue("
Based on {baseline$conv_pct}% MCMC convergence and a mean posterior SD of
**{baseline$phi_main_se}** for mainstem survival and **{baseline$psi_down_se}**
for downstream emigration probability at the 1,000-fish baseline scenario,
the following recommendations apply:
"))
}
```

Regardless of simulation outcome, the following structural conclusions hold:

**Minimum recommended tagging effort: 1,000 fish**
: Below this threshold, emigration estimates become unstable due to the small
  number of fish passing antennas. At 11–15% emigration, only 110–150 emigrants
  are expected per 1,000 tagged fish — the lower bound of what the Gembeck/Sager
  double-observer system needs to reliably estimate detection efficiency.

**Target tagging effort: 2,000–4,000 fish**
: Achievable within the spring electrofishing + screw trap design (pilot estimate:
  4,000–6,000 fish). This range meaningfully improves precision over the 1,000-fish
  minimum and provides robustness against poor electrofishing years (p_efish ≈ 0.15).

**Value of pilot detection data**
: Informative priors on detection probabilities (from the 2023–2025 pilot work)
  materially improve estimation of downstream emigration and Gembeck detection
  efficiency. The pilot data are load-bearing — vague priors produce noticeably
  wider posteriors, particularly when survival is low and few fish are available
  at each antenna.

**Primary study reach: Warm Springs to Gembeck Rd**
: The ~11-mile reach between these two mainstem antennas is appropriately sized
  for boat electrofishing and provides a bounded study area for survival estimation.
  Extension to the Gembeck–Sager section is worthwhile if tagging targets are met,
  but the upper reach should be the analytical priority.

---

## Methods Code

The full simulation and model-fitting code is available in this repository.
Simulation is deterministic given the scenario inputs; results reproduce exactly
with `set.seed(123)`.

```{r}
#| label: show-sim-code
#| eval: false
#| echo: true
#| code-fold: false
#| file: power_analysis_CFBrown_loop.R
```

---

*Analysis conducted in R using `jagsUI` for Bayesian MCMC estimation.
Model: `survival_CFBrown.txt`. Rendered with [Quarto](https://quarto.org).*
